---
title: "Writing an adapter"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Writing an adapter}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(linop)
set.seed(4)
```

If you have a class that represents a linear map, one method makes it work
everywhere a `linop` works: in the algebra, in `verify()`, and in the solvers
that arrive with the next phase. This vignette writes that method for a real
operator, reads the certificate line by line, and then does the same job the
other way, by registering a node type.

## The example

A circulant matrix is determined by its first column, and it is applied by
pointwise multiplication in the Fourier domain. An n by n circulant costs
`O(n log n)` per apply and `O(n)` to store, so it is exactly the sort of operator
that should never be written out.

```{r conv}
conv <- function(w, X) {
  n <- length(w)
  Y <- X
  for (j in seq_len(ncol(X))) {
    z <- fft(w * fft(X[, j]), inverse = TRUE) / n
    Y[, j] <- if (is.complex(X)) z else Re(z)
  }
  Y
}
```

`w` is the symbol, `fft(col1)`. Two details in that function are the whole
contract in miniature. The loop is over the columns of a block, because an apply
always receives a matrix. The last line preserves the input type: a real block
gives a real result, a complex block keeps its imaginary part.

## One method

`linop()` is a generic dispatching on its first argument. An adapter defines
`linop.myclass()` and returns a `linop`.

```{r adapter}
circulant <- function(col1) structure(list(col1 = col1), class = "circulant")

linop.circulant <- function(x, ...) {
  n <- length(x$col1)
  w <- fft(x$col1)
  linop(
    function(X) conv(w, X),
    adjoint = function(X) conv(Conj(w), X),
    dim = c(n, n),
    cost = 15 * n * log2(n)
  )
}
```

In a package, that method is registered with `S3method(linop, circulant)` in
`NAMESPACE`, which `roxygen2` writes from an `@export` tag on the method.

That is the whole adapter. The object now behaves like a matrix:

```{r use}
C <- linop(circulant(c(1, 2, 3, 0, 0, 0)))
C
zapsmall(C %*% c(1, 0, 0, 0, 0, 0))
```

and it composes:

```{r compose}
N <- crossprod(C) + 0.1 * linop_eye(6)
N
cap(N, "hermitian")$value
```

The `cost` argument is worth passing. Without it a callback leaf assumes the
dense flop count, `2 * nrow * ncol`, which would make a fast operator look
expensive to `collapse()` and to any future planner.

```{r cost}
linop_cost(C)
2 * 6 * 6      # what the default would have been
```

## Reading the certificate

```{r verify}
verify(C)
```

Eleven checks, each reporting its own status, its source and the kind of
guarantee it gives.

**adjoint consistency** tests `<y, A x> = <A^H y, x>` over random probe pairs.
This is the check that earns its keep. A missing or misplaced conjugation is the
most common bug in a hand-written operator, it produces an object that looks
right under every casual test, and it makes every Krylov method built on it
converge to the wrong answer. Here is the same adapter with `Conj()` dropped:

```{r verify-bad}
w <- fft(c(1, 2, 3, 0, 0, 0))
Bad <- linop(function(X) conv(w, X),
             adjoint = function(X) conv(w, X),   # missing the conjugation
             dim = c(6, 6))
verify(Bad)
```

**view identities** checks that `adjoint(t(A))`, `t(adjoint(A))` and
`adjoint(Conj(A))` act as they must. **real transpose equals adjoint** appears
only for a real operator, and asserts both that the two act alike and that they
remain distinct nodes.

**linearity** tests `A(a x1 + b x2) = a A x1 + b A x2` with complex coefficients
when the operator is complex. It is also what catches an adaptive solve that has
been wrapped as an operator.

**block consistency** applies a block of several columns and compares each output
column against applying that column alone. An operator that treats its input as a
vector passes every scalar test and fails here.

**shape and dtype** applies real and complex blocks and checks the result shape,
and that a complex input produces a complex output.

**complex linearity** tests `A(x + iy) = A(x) + i A(y)`. This is the check that
`conv()` was written to satisfy, in the line that keeps a complex block complex.
An apply that discards the imaginary part of its input passes every other check
in the suite, because the apply path promotes the real result it returned to
complex before handing it back, which is enough to satisfy `shape and dtype`.

**declared capabilities** probes anything the operator claims about itself.
Agreement is recorded as a probe beside the declaration; a contradiction fails
the certificate. Nothing was declared above, so the line reads `not_checked`.

**purity** applies the same block twice and requires identical output and an
unmodified input.

**gemm agreement** checks the two-tier apply. Authors implement tier one,
`apply(op, X, mode)`; solvers only ever call tier two,
`alpha * op(A) X + beta * Y`, which is synthesised from tier one. This line
confirms the synthesis matches.

**materialisation agreement** compares the dense form against applying the
operator to the columns of an identity, when the operator is small enough to
materialise at all.

The overall verdict is `pass`, `qualified` or `fail`. `qualified` means every
check that ran passed and at least one did not run, so the certificate above is
qualified only because nothing was declared to probe.

```{r cert-fields}
cert <- verify(C)
cert$overall
cert$without_deterministic_bound
```

The certificate is a list, so a package can test against it directly:

```{r cert-test}
stopifnot(verify(C)$overall %in% c("pass", "qualified"))
```

## Declaring what you know

If the adapter knows a property, it should say so, and then `verify()` will probe
it.

```{r declared}
ws <- fft(c(2, -1, 0, 0, 0, -1))      # circularly symmetric kernel

Sym <- linop(function(X) conv(ws, X),
             adjoint = function(X) conv(Conj(ws), X),
             dim = c(6, 6),
             properties = c(symmetric = TRUE))
verify(Sym)$probes
```

One declaration, two probe rows. A real symmetric operator is Hermitian by
definition, so the closure applied at construction added the second claim and
`verify()` probed both.

A declaration the operator does not honour is a failure rather than a warning:

```{r declared-bad}
wa <- fft(c(1, 2, 3, 0, 0, 0))        # the asymmetric kernel from above

Liar <- linop(function(X) conv(wa, X),
              adjoint = function(X) conv(Conj(wa), X),
              dim = c(6, 6),
              properties = c(symmetric = TRUE))
verify(Liar)$probes
verify(Liar)$overall
```

## Registering a node type

The callback form covers most adapters. Storage that wants its own place in the
expression graph, its own cost function and its own materialisation registers a
node type instead. The internal nodes go through this same door, which is what
keeps it honest.

A node handler is three functions. `apply(op, X, mode)` is the only required one.

```{r register}
flip <- function(w) w[c(1L, length(w):2L)]

linop_register_node("circulant",
  apply = function(op, X, mode) {
    w <- op$args$symbol
    conv(switch(mode, N = w, T = flip(w), C = Conj(w), R = Conj(flip(w))), X)
  },
  materialize = function(op) {
    E <- diag(1, op$dim[1L])
    if (op$dtype == "complex") storage.mode(E) <- "complex"
    conv(op$args$symbol, E)
  },
  cost = function(op) 15 * op$dim[1L] * log2(op$dim[1L]))

linop_nodes()
```

The four modes are the BLAS convention: `N` is `A X`, `T` is `A^T X`, `C` is
`A^H X`, `R` is `conj(A) X`. A callback leaf supplies two of them and lets the
package derive the rest; a node handler is asked for all four, and in exchange
gets to implement each one at its natural cost. For a circulant, the symbol of
the transpose is the cyclic reversal of the symbol and the symbol of the adjoint
is its conjugate, so all four are one multiply.

Nodes are constructed through the package's internal constructor, so in practice
a node type ships with a user-facing constructor beside it:

```{r node-construct}
circ_node <- function(col1) {
  n <- length(col1)
  linop:::new_linop("circulant", c(n, n),
                    dtype = if (is.complex(col1)) "complex" else "double",
                    args = list(symbol = fft(col1)))
}

Cn <- circ_node(c(1, 2, 3, 0, 0, 0))
Cn
linop_cost(Cn)
```

Run the same conformance suite. It is the same suite, and it is where a mode you
got wrong shows up:

```{r node-verify}
verify(Cn)$overall
```

Complex kernels exercise all four modes independently, which is the case worth
testing:

```{r node-complex}
Zn <- circ_node(c(1 + 1i, 2, 0, 0, 3 - 1i, 0))
verify(Zn)$overall
```

The `materialize` handler above is written the way it is for a reason found by
running exactly this check. The first version built `diag(1, n)`, a real
identity, and `conv()` faithfully preserved the input type by discarding the
imaginary part of a complex operator's own columns. Every other line of the
certificate passed and `materialisation agreement` failed with a gap of 1.0.

## Provenance

An operator that came from somewhere else can carry an envelope back to its
provider. The core stores it and never looks inside, which keeps concepts like
discretisation, lifting and refinement out of the object model while still
letting a provider recover them.

```{r provenance}
Cp <- set_provenance(Cn, provider = "mypkg",
                     payload = list(h = 0.1, grid = "uniform", n_original = 6))

provenance(Cp)$provider
provenance(Cp)$payload$h
```

Four generics are dispatched on the envelope, and a provider implements the ones
it can support: `provenance_lift()` maps a finite vector into the provider's
space, `provenance_refine()` produces a finer discretisation,
`provenance_original_residual()` measures a finite result against the original
problem, and `provenance_summary()` describes the envelope in one line. The
default methods raise an error naming the provider that did not register one,
apart from `provenance_summary()`, which degrades to a description.

```{r provenance-summary}
provenance_summary(provenance(Cp))
```

The envelope travels through views, scales, sums and products, and
`strip_provenance()` removes it.

```{r provenance-travel}
provenance(2 * Cp)$provider
is.null(provenance(strip_provenance(Cp)))
```

## Preconditioners carry a contract

Two representations of a preconditioner are common, and mixing them is a
recurring source of bugs: `M` approximating `A` and applied by solving
`M z = r`, and `P` approximating `A^-1` and applied as `z = P r`. Only one
direction exists internally, `apply_inverse`, and the two constructors say which
one you have.

```{r precond}
D <- linop_scaling(c(4, 9, 16))

P1 <- as_preconditioner(D, solver = function(M, R) R / M$args$d)   # M ~ A
P2 <- as_preconditioner_inverse(linop_scaling(1 / c(4, 9, 16)))    # P ~ A^-1
P2
```

Both constructors read `hermitian` and `positive_definite` off the operator they
were given, so a preconditioner built from a `linop` inherits what that operator
knows.

Methods declare what they require, and a preconditioner that does not meet the
requirement is refused by name. A flexible preconditioner, one whose map changes
between applications, breaks the short recurrence that conjugate gradients is
built on:

```{r precond-flexible}
flexible <- preconditioner(function(R) R, fixed = FALSE)
flexible$fixed
```

Passing that to CG raises an error naming `fixed` and pointing at `fgmres`, which
is the method designed to accept it. The requirement table covers `cg`, `minres`,
`gmres`, `fgmres`, `bicgstab`, `lsqr` and `lsmr`, and it is enforced at the point
the preconditioner is accepted rather than at the point the iteration misbehaves.

## Solve objects are not operators

One boundary is worth knowing before the solvers land. An object is an operator
if and only if its action on `x` is determined without reference to `x`.
Conjugate gradients chooses its polynomial by minimising over the Krylov space of
the right-hand side, so an approximate solve is not a linear operator, and it is
not a `linop`.

Approximate solves are a separate class carrying a declared contract: whether the
solve is exact, a fixed linear approximation, or variably inexact; whether it is
fixed, input-dependent, or history-dependent; whether it is deterministic;
whether an adjoint and an error bound are available. Consumers declare which
contracts they accept, so an ordinary Arnoldi iteration cannot silently receive a
history-dependent solve.

`verify()` dispatches on that class and tests the declared contract instead of
assuming purity, and running the operator suite against one is an error that says
why. The class is internal in this release; `solve()` is the public front door
arriving with the next phase.

## A checklist

Before shipping an adapter:

1. `verify()` returns `pass` or `qualified`, on a real instance and, if the
   format supports it, a complex one.
2. `qualified` is explained. Which check did not run, and why.
3. Declared capabilities are honoured, and `verify(x)$probes` says
   `consistent`.
4. `cost` is passed, so the operator is not mistaken for a dense matrix.
5. Blocks of width greater than one behave, which `block consistency` covers.
6. The apply does not modify its input, which `purity` covers.

## Where to next

- `vignette("capabilities")` covers what to declare and what a solver will do
  with it.
- `vignette("expressions")` covers the cost model your `cost` function feeds.
