---
title: "Capabilities and evidence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Capabilities and evidence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(linop)
set.seed(3)
```

A solver needs to know things about an operator that it cannot afford to find
out. Conjugate gradients converges on a Hermitian positive definite operator and
can silently produce nonsense on one that is neither, and checking positive
definiteness costs more than the solve. So the operator has to carry the claim.

The moment it carries a claim, a second question arrives: on whose word. A
property established by construction, one measured exactly on stored data, one
asserted by whoever wrote the adapter, and one guessed from ten random probes are
all "positive definite is TRUE", and they are not interchangeable. This vignette
is about keeping the two apart and keeping both correct through the algebra.

## Ten capabilities, three values

```{r names}
caps <- c("hermitian", "symmetric", "positive_definite", "positive_semidefinite",
          "unitary", "diagonal", "triangular_upper", "triangular_lower",
          "full_rank", "real")
```

`hermitian` and `symmetric` are separate entries. For a complex operator
`A = A^H` and `A = A^T` are different statements, and a solver that takes one for
the other takes a wrong branch. For a real operator they coincide, and the
package records that coincidence as a rule instead of merging the fields.

Each capability holds `TRUE`, `FALSE` or `NA`. `NA` means unknown, and it is
never read as `FALSE`.

```{r three-valued}
S <- linop(function(X) X[c(2, 1), , drop = FALSE],
           adjoint = function(X) X[c(2, 1), , drop = FALSE],
           dim = c(2, 2))
cap(S, "positive_definite")
```

That operator is a swap. It is Hermitian and it is not positive definite, and
`linop` says neither, because nothing has established either one. Propagation is
conservative in the same direction: a rule that cannot prove `TRUE` returns `NA`
and never `FALSE`.

`cap()` reads one entry. `summary()` reads them all:

```{r summary}
A <- linop(matrix(c(2, 1, 0,
                    1, 3, 1,
                    0, 1, 2), 3, 3))
summary(A)$caps[, c("capability", "value")]
```

The dense leaf filled those in by checking the stored matrix exactly. Symmetry
came from `all.equal(M, t(M))`, diagonality from testing the off-diagonal
entries. Positive definiteness is absent because establishing it would have cost
a factorisation the constructor did not run.

## Evidence is three independent fields

```{r evidence-fields}
evidence(source = "computation", guarantee = "identity", confidence = 1)
```

**source** is where the claim came from: `construction`, `adapter_contract`,
`user_declaration`, `probe`, `computation`, `theorem`.

**guarantee** is the kind of statement being made: `identity`,
`deterministic_bound`, `probabilistic_bound`, `estimate`, `heuristic`.

**confidence** is `1`, a stated probability, or `NA`.

The three are independent on purpose, and none of them is a rank. A randomised
trace estimator carries a `probabilistic_bound` with a rigorous confidence; a
deterministic heuristic carries no bound at all; an adapter contract can be worth
more than a hundred probes. Collapsing that into one "strength" number loses
exactly the distinctions a solver needs.

The fourth field is what makes it work: `depends_on`, a list of the evidence a
rule relied on. It is empty when the rule is unconditional.

## Declaring what you know

`properties` takes a named logical vector and records each entry as a user
declaration.

```{r declare}
P <- linop(function(X) X, adjoint = function(X) X, dim = c(4, 4),
           properties = c(hermitian = TRUE, positive_definite = TRUE))
cap(P, "positive_definite")
```

A bare character vector is shorthand for all-`TRUE`, and an unknown name is an
error naming the ten that exist.

```{r declare-bad, error = TRUE}
linop(matrix(1:4, 2, 2), properties = c(hermitiann = TRUE))
```

Some implications hold by definition and are closed once at construction, so
declaring the strong form does not oblige you to also declare the weak one:

```{r closure}
cap(P, "positive_semidefinite")
```

Positive definite implies positive semidefinite; diagonal implies both
triangularities; a real Hermitian operator is symmetric and the converse. Each
closure records what it depended on. A `"double"` operator is `real` by
construction, unconditionally.

```{r closure-real}
cap(linop(matrix(1:4, 2, 2)), "real")
```

Declaring `hermitian` or `symmetric` on a non-square operator is an error rather
than a stored contradiction:

```{r declare-nonsquare, error = TRUE}
linop(matrix(1:6, 2, 3), properties = c(symmetric = TRUE))
```

## Propagation

Every composite recomputes its capabilities from its operands. The rules are
sound in the direction that matters: they claim `TRUE` only where the conclusion
follows, and answer `NA` everywhere else.

**Sums.** `hermitian`, `symmetric`, `positive_definite`, `positive_semidefinite`,
`real`, `diagonal` and both triangularities survive addition when every term has
them.

```{r prop-sum}
D1 <- linop_scaling(c(1, 2, 3))
D2 <- linop_scaling(c(4, 5, 6))
cap(D1 + D2, "diagonal")$value
cap(D1 + D2, "positive_definite")$value
```

**Scaling.** A real scalar preserves `hermitian`; any scalar preserves
`symmetric`; a positive real scalar preserves definiteness. A nonzero scalar
preserves `full_rank`, and structure such as `diagonal` survives any scalar.

```{r prop-scale}
Pd <- linop(function(X) X, adjoint = function(X) X, dim = c(3, 3),
            properties = c(hermitian = TRUE, positive_definite = TRUE))

cap(2 * Pd, "positive_definite")$value
cap(-2 * Pd, "positive_definite")$value    # the rule cannot conclude
cap(-2 * Pd, "hermitian")$value            # a real scalar keeps this one
```

A diagonal leaf answers `FALSE` to the same question, and the difference is
worth understanding. Scaling a `diag` is a simplification rule: the result is a
fresh diagonal leaf, whose constructor computes definiteness exactly from the
stored vector rather than inheriting a propagated claim.

```{r prop-scale-diag}
cap(-2 * D1, "positive_definite")$value
(-2 * D1)$node
```

**Transpose, adjoint, conjugate.** `adjoint()` preserves `hermitian`,
definiteness, `unitary`, `diagonal`, `full_rank` and `real`, and preserves
`symmetric` when the operand is also real. `t()` preserves `symmetric` and
preserves `hermitian` when the operand is real. Both flip the triangularities.
`Conj()` preserves everything.

```{r prop-views}
U <- linop(matrix(c(1, 0, 2, 1), 2, 2))      # upper triangular
cap(U, "triangular_upper")$value
cap(t(U), "triangular_upper")$value
cap(t(U), "triangular_lower")$value
```

**Products.** A product of two arbitrary operators tells you nothing, and the
rules say so. Two exceptions are structural, and they hold whatever is known
about the operand:

```{r prop-gram}
K <- linop(matrix(rnorm(12), 4, 3))
G <- adjoint(K) %*% K
cap(G, "hermitian")
cap(G, "positive_semidefinite")$value
cap(G, "positive_definite")$value            # NA until K is known full rank
```

`adjoint(X) %*% X` is Hermitian and positive semidefinite for every `X`, so the
evidence has nothing underneath it. Positive definiteness needs `X` to have full
rank, so it appears only when that is known, and then it records the dependency.

The product node finds this by structural recognition, comparing the two factors
rather than testing the result. It is conservative: it may fail to notice that
two separately built operators are equal, and it never claims that two different
operators are the same.

```{r prop-gram-real}
Kr <- linop(matrix(rnorm(12), 4, 3))
cap(crossprod(Kr), "symmetric")$value        # t(A) A is symmetric for any A
cap(crossprod(Kr), "hermitian")$value        # and Hermitian too, since A is real
```

For a complex operator the two come apart, which is the point of keeping them
separate:

```{r prop-gram-complex}
Kz <- linop(matrix(c(1 + 1i, 2, 0, 3 - 2i), 2, 2))
cap(crossprod(Kz), "symmetric")$value        # A^T A is symmetric
cap(crossprod(Kz), "hermitian")$value        # and is not Hermitian
cap(adjoint(Kz) %*% Kz, "hermitian")$value   # A^H A is
```

Structure that survives any product is carried through: a product of diagonals is
diagonal, of upper triangulars is upper triangular, of unitaries is unitary, of
reals is real.

```{r prop-structure}
cap(D1 %*% D2, "diagonal")$value
```

## The case this design exists for

Take two operators whose author declared them positive definite. The sum rule is
unconditionally sound. So the sum is positive definite, by construction:

```{r launder}
P <- linop(function(X) X, adjoint = function(X) X, dim = c(4, 4),
           properties = c(hermitian = TRUE, positive_definite = TRUE))
Q <- linop(function(X) 2 * X, adjoint = function(X) 2 * X, dim = c(4, 4),
           properties = c(hermitian = TRUE, positive_definite = TRUE))

cap(P + Q, "positive_definite")
```

Read the evidence. The source is `construction`, which is true and is not the
whole story: underneath it sit the two user declarations the rule consumed. Had
the composite reported a bare `construction`, two unverified assertions would
have been laundered into a fact by one addition, and a solver reading only the
top-level source would accept them.

## Requirements

A solver states what it will accept. The statement is a predicate over evidence,
not a threshold on a strength number.

```{r requirement}
strict <- requirement(sources = c("construction", "computation", "theorem"),
                      guarantees = "identity")
```

`evidence_satisfies()` checks the evidence against it, and recurses into
`depends_on`, so a requirement the inputs would have failed is failed by anything
built from them.

```{r satisfies}
evidence_satisfies(cap(P, "positive_definite")$evidence, strict)
evidence_satisfies(cap(P + Q, "positive_definite")$evidence, strict)
```

The Gram matrix passes the same requirement, because there is nothing underneath
its evidence to fail it:

```{r satisfies-gram}
evidence_satisfies(cap(G, "positive_semidefinite")$evidence, strict)
```

A solver willing to take the author's word says so explicitly, and then the same
chain passes:

```{r satisfies-lenient}
lenient <- requirement(sources = c("construction", "computation", "theorem",
                                   "user_declaration"))
evidence_satisfies(cap(P + Q, "positive_definite")$evidence, lenient)
```

`min_confidence` adds a numeric floor, applied at every level of the chain:

```{r min-conf}
probabilistic <- evidence("probe", "probabilistic_bound", confidence = 0.99)
evidence_satisfies(probabilistic, requirement(min_confidence = 0.999))
evidence_satisfies(probabilistic, requirement(min_confidence = 0.95))
```

This is the shape of the contract Phase 2 solvers will use: CG will require
Hermitian positive definiteness under a requirement it publishes, and an operator
whose claim rests on an unverified declaration will be turned away by name rather
than converge to something wrong.

## Probing a declaration

`verify()` tests declared capabilities against the operator's actual behaviour.
Agreement is recorded beside the declaration as a probe and never promoted to
proof; a contradiction fails the certificate.

```{r probe}
Good <- linop(function(X) X, adjoint = function(X) X, dim = c(4, 4),
              properties = c(hermitian = TRUE))
verify(Good)$probes
```

```{r probe-bad}
Liar <- linop(function(X) rbind(X[-1, , drop = FALSE], 0),
              adjoint = function(X) rbind(0, X[-4, , drop = FALSE]),
              dim = c(4, 4), properties = c(hermitian = TRUE))
verify(Liar)$probes
verify(Liar)$overall
```

Positive definiteness is the one case where a probe can only ever disagree. A
positive Rayleigh quotient on ten random vectors is not evidence of definiteness,
so the check looks for a negative one and reports `consistent` otherwise. The
certificate records the guarantee as `heuristic` for exactly this reason.

## Where to next

- `vignette("adapters")` covers the rest of the certificate, and the contracts
  that preconditioners and solve objects carry.
- `vignette("expressions")` covers the tree these capabilities travel up.
- `vignette("solvers")` covers what `method = "auto"` does with the evidence
  above, which is where these three fields decide something visible.
- `vignette("certificates")` covers the same three fields underneath a computed
  result, including a bound that inherits the evidence of the capability it
  rests on.
