---
title: "Reading a certificate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading a certificate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(linop)
set.seed(1)
```

## One object, four shapes

Every computation in this package that could be wrong comes back with a
certificate. It is the same class each time, and it takes four shapes, because
the questions differ:

```{r shapes, echo = FALSE}
knitr::kable(data.frame(
  produced_by = c("verify()", "solve(), square", "solve(), least squares", "eigs(), svds()"),
  subject = c("does the operator behave like a linear map",
              "how close is x to solving A x = b",
              "how close is x to minimising ||b - A x||",
              "how close is (theta, x) to an eigenpair"),
  check.names = FALSE))
```

This article is about the last three. `verify()` is covered in
`vignette("adapters")`, where it belongs to the job of writing an operator.

## Status and evidence are separate questions

Every row carries a status and, separately, the kind of argument behind it.

```{r first}
n <- 200
lap <- function(X) 2 * X - rbind(X[-1, , drop = FALSE], 0) - rbind(0, X[-n, , drop = FALSE])
L <- linop(lap, adjoint = lap, dim = c(n, n), properties = c(hermitian = TRUE))
b <- as.numeric(L %*% sin(seq_len(n) * pi / (n + 1)))

cert <- attr(solve(L, b, method = "minres", tol = 1e-10), "certificate")
cert
```

The status column answers whether the check passed. `pass`, `fail`,
`not_checked`, and `qualified`, which means it passed for a reason weaker than
the clean one.

The three columns after it are the evidence, and they are three independent
fields rather than a ranking. `source` is what kind of argument produced the
number: a `computation` on data at hand, a `theorem`, a `construction`, a
`user_declaration`. `guarantee` is what the argument gives you: an `identity`
holds exactly, a `deterministic_bound` holds always, an `estimate` holds usually.
`confidence` is a probability where one applies and `-` where none does.

A `pass` behind an `estimate` and a `pass` behind an `identity` are both passes
and they are not the same claim. The summary line names the rows with neither
kind of deterministic guarantee, so that distinction survives being skimmed:

```{r weak}
cert$without_deterministic_bound
```

## The rows of a linear solve

```{r rows-square, echo = FALSE}
knitr::kable(data.frame(
  row = c("arithmetic floor", "residual", "backward error", "convergence",
          "forward error"),
  measures = c("the level below which no residual is meaningful",
               "||b - A x|| / ||b||",
               "||r|| / (||A|| ||x|| + ||b||), the Rigal-Gaches quantity",
               "whether the iteration stopped because it was done",
               "how far x is from the true solution"),
  check.names = FALSE))
```

The backward error is the row to read. It answers a question that has an exact
answer: it is the size of the smallest perturbation to `A` and `b` for which your
computed `x` is the exact solution. A backward error of 1e-14 means you solved a
problem within 1e-14 of the one you asked about.

`||A||` in that denominator is a lower bound rather than an estimate in either
direction, deliberately. Shrinking the denominator can only make the reported
backward error larger, so the number can overstate the error and never understate
it.

## The arithmetic floor

Ask for an exact solve and no floating-point iteration can deliver it:

```{r floor}
tight <- solve(L, b, method = "minres", tol = 0, maxit = 400)
bare  <- solve(L, b, method = "minres", tol = 0, maxit = 400,
               control = list(floor_const = 0))

identical(as.numeric(tight), as.numeric(bare))
```

The same iteration ran twice and produced the same vector, bit for bit. Only the
certificates differ:

```{r floor-certs}
c(with_floor = attr(tight, "certificate")$overall,
  without    = attr(bare,  "certificate")$overall)
```

```{r floor-detail}
attr(bare, "certificate")
```

A backward error of 1.8e-16 is reported as a failure, because it was tested
against a tolerance of zero with nothing underneath it. The floor is what stops a
fully converged solve from certifying as `fail`, and it is
`c * eps * (||A|| ||x|| + ||b||)`, which is Higham's bound on the computed
residual. That is the same quantity as the Rigal-Gaches denominator, so once the
backward error is written relatively the floor is a plain `c * eps`.

A row that meets its tolerance only because of the floor says `qualified` and
carries an `estimate`, never a clean `identity`. `c` counts terms in an inner
product a matrix-free operator never exposes, so it is an argument with a default
rather than a constant.

## The same table, measured differently

A least-squares solve has no exact answer. `b - A x` does not go to zero, so
testing it against a tolerance would report a converged solution as a failure.
What goes to zero is `A^H r`.

Both are backward errors, so one row carries both readings, and which one applies
is decided by measurement rather than by the method that was called:

```{r lsq}
G <- matrix(rnorm(120 * 40), 120, 40)
R <- linop(G)

incompatible <- solve(R, rnorm(120), method = "lsmr", tol = 1e-10)
compatible   <- solve(R, as.numeric(G %*% rnorm(40)), method = "lsmr", tol = 1e-10)

status_of <- function(x, row) {
  ch <- attr(x, "certificate")$checks
  ch$status[ch$check == row]
}
c(incompatible = status_of(incompatible, "residual"),
  compatible   = status_of(compatible,   "residual"))
```

For the incompatible right-hand side the residual reads `not_checked`, because it
has a nonzero minimum and comparing that to `tol` would mean nothing. For the
compatible one the same code found a residual that did go to zero and checked it.

```{r lsq-cert}
attr(incompatible, "certificate")
```

The perturbation behind the least-squares reading is Stewart's,
`dA = -(r r^H A) / ||r||^2`, which is exhibited rather than argued: it is an
actual matrix of that norm making `x` an exact minimiser. That is what makes
`||A^H r|| / (||A|| ||r||)` a backward error rather than a heuristic, and it is
also Paige and Saunders' stopping rule.

## A third shape, for an eigenpair

For a solve, `b` is data and `A` is data, and the Rigal-Gaches denominator has a
term for each. For an eigenpair, `theta` and `x` are both outputs and `A` is the
only datum, so the denominator loses a term and the table gains rows:

```{r eigen}
fit <- eigs(L, k = 3, which = "largest_algebraic", ncv = 60)
fit$certificate
```

`orthogonality` is new: the computed vectors should be orthonormal, and this
measures whether they are. `target identity` is new and is always `not_checked`,
which is the honest entry. A small residual proves the pair is an approximate
eigenpair. That it is the *largest* one needs an inertia count, an enclosure or a
separation bound, and matrix-free there is generally none available.

## The one row that is sometimes a bound

`forward error` is `not_checked` on every linear solve in this article, and it is
not laziness. Nothing constrains it. At a condition number of 1e10:

```{r forward}
m <- 40
Q <- qr.Q(qr(matrix(rnorm(m * m), m, m)))
M <- Q %*% diag(10^seq(0, -10, length.out = m)) %*% t(Q)
M <- (M + t(M)) / 2
Aill <- linop(M)
xtrue <- rnorm(m)
bill <- as.numeric(Aill %*% xtrue)

t(vapply(c("minres", "gmres", "bicgstab"), function(mth) {
  z <- solve(Aill, bill, method = mth, tol = 1e-10, maxit = 400)
  c(backward = attr(z, "certificate")$values$backward_error,
    forward  = max(abs(z - xtrue)) / max(abs(xtrue)))
}, numeric(2)))
```

Backward errors around 1e-11 sit beside true forward errors around 1e-1. Both
columns are correct. `kappa^2 * eps` is 2.2e4 here, and no backward error
constrains the forward one at that conditioning, which is exactly why the row
refuses to report a number it cannot stand behind. Do not read that spread as one
method being more accurate than another.

For an eigenpair the same row does carry a bound, and it is the first in the
package. `A + E` with `E = -r x^H - x r^H` is hermitian and has `theta` as an
exact eigenvalue, so Weyl's inequality gives
`min_j |theta - lambda_j(A)| <= ||r|| / ||x||`:

```{r forward-eigen}
fwd <- fit$certificate$checks
fwd <- fwd[fwd$check == "forward error", ]
fwd[, c("status", "source", "guarantee")]
cat(fwd$detail)
```

`source` is `theorem` and `guarantee` is `deterministic_bound`, the only place
either appears. The bound is about *some* eigenvalue of `A`; which one is the
`target identity` question and stays `not_checked`.

## Evidence can carry a dependency

That bound holds because `A` is hermitian. If the only argument for that is your
say-so, the bound inherits it, and the certificate records the inheritance rather
than hiding it:

```{r depends}
Md <- crossprod(matrix(rnorm(20 * 20), 20, 20)) + diag(20)

checked  <- eigs(linop(Md), k = 2, which = "largest_algebraic")
declared <- eigs(linop(function(X) Md %*% X, adjoint = function(X) Md %*% X,
                       dim = c(20, 20), properties = c(hermitian = TRUE)),
                 k = 2, which = "largest_algebraic")

checked$certificate$evidence[["forward error"]]
declared$certificate$evidence[["forward error"]]
```

Both say `theorem / deterministic_bound`. What differs is in the brackets, and it
is what a requirement tests:

```{r depends-req}
strict <- requirement(sources = c("construction", "computation", "theorem"),
                      guarantees = c("identity", "deterministic_bound"))

c(checked  = evidence_satisfies(checked$certificate$evidence[["forward error"]],  strict),
  declared = evidence_satisfies(declared$certificate$evidence[["forward error"]], strict))
```

A theorem applied to a declaration is worth what the declaration is worth. This
is the same rule that stops a sum of two declared positive-definite operators
from satisfying CG's requirement, described in `vignette("capabilities")`,
reaching the certificate.

## Reading one in code

The printed table is for people. The object is a list, so a package can test
against it:

```{r programmatic}
names(cert)
cert$checks[, c("check", "status")]
cert$values$backward_error
```

```{r assert}
stopifnot(cert$overall %in% c("pass", "qualified"))
```

`cert$dispatch` records which method ran and why, `cert$values` holds the
measured quantities, and `cert$evidence` holds the evidence objects for rows that
carry a dependency.

## Where to next

- `vignette("solvers")` covers the methods that produce the first two shapes.
- `vignette("spectral")` covers `eigs()` and `svds()`, including a result whose
  values are exact and whose certificate reads `fail`.
- `vignette("capabilities")` covers evidence away from certificates, where the
  same three fields decide what a solver will accept.
- `vignette("adapters")` covers `verify()`, the fourth shape.
