---
title: "The expression graph"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The expression graph}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(linop)
set.seed(2)
```

Every arithmetic operation on a `linop` returns a new operator holding its
operands. The result is a tree, and everything the package does afterwards is a
walk over that tree: applying, printing, costing, materialising, propagating what
is known. This vignette is about the tree.

## Leaves and composites

```{r nodes}
linop_nodes()
```

Five of those are leaves and hold data or a callback:

| node | holds | applied as |
|---|---|---|
| `dense` | a base matrix | `M %*% X` |
| `sparse` | a `Matrix` | `M %*% X`, real storage only |
| `fun` | one or two closures | `f(X)` |
| `identity` | a size | `X` |
| `diag` | a vector | `d * X` |

Six are composites and hold other operators:

| node | built by | applied as |
|---|---|---|
| `transpose` | `t()` | the operand under mode `T` |
| `adjoint` | `adjoint()` | the operand under mode `C` |
| `conjugate` | `Conj()` | the operand under mode `R` |
| `scale` | `a * A` | `a` times the operand |
| `sum` | `A + B` | the operands added |
| `product` | `A %*% B` | the operands applied right to left |

The list is closed on purpose for this release. Low-rank, Kronecker, stacked,
block-diagonal, permutation, power and inverse nodes are deferred until an
external adapter has exercised the registry, and a test asserts they are absent
so that the deferral is a mechanism rather than an intention.

## What the tree looks like

```{r tree}
n <- 6
S <- linop(function(X) rbind(X[-1, , drop = FALSE], 0),
           adjoint = function(X) rbind(0, X[-n, , drop = FALSE]),
           dim = c(n, n))

E <- crossprod(S) + 0.1 * linop_eye(n)
E
```

`explain()` gives the same tree as a table, with the estimated cost of each node:

```{r explain}
explain(E)
```

Depth zero is the root. The `product` under it came from `crossprod`, which is
`t(S) %*% S`, so its first child is a `transpose` view over the same `fun` leaf
that appears again as its second child. The leaf is shared, not copied: a `linop`
is immutable, so an operator that appears twice in an expression is the same
object twice.

## Simplification happens at construction

A handful of rewrites are applied as the expression is built. They are local,
they never consult a cost model, and they exist so that the tree stays small and
the obvious identities hold on the nose.

Views involute and compose:

```{r simplify-views}
Rect <- linop(matrix(1:6, 2, 3))
t(t(Rect))$node
adjoint(adjoint(Rect))$node
Conj(Conj(Rect))$node
adjoint(t(Rect))$node
Conj(adjoint(Rect))$node
```

Scales fold together, and a scale of a diagonal is a diagonal:

```{r simplify-scale}
D <- linop_scaling(c(1, 2, 3))
(2 * (3 * D))$node
(2 * (3 * D))$args$d
```

Sums flatten and drop zero terms, products flatten and drop identities:

```{r simplify-sum}
A <- linop(matrix(rnorm(9), 3, 3))
length(((A + D) + A)$args$ops)
(A %*% linop_eye(3))$node
```

Adjacent diagonals in a product fuse into one:

```{r simplify-diag}
D2 <- linop_scaling(c(4, 5, 6))
(D %*% D2)$node
(D %*% D2)$args$d
```

Diagonal and identity leaves absorb a view outright, since transposing them is
either nothing or a conjugation of the stored vector:

```{r simplify-absorb}
Dz <- linop_scaling(c(1 + 1i, 2))
t(Dz)$node
adjoint(Dz)$node
adjoint(Dz)$args$d
```

One rewrite that deliberately does **not** happen: for a real operator, `t()` and
`adjoint()` have identical action, and neither is rewritten into the other.
Keeping them distinct is what lets `print()` show the expression as written, and
it keeps the distinction alive at the boundary where a real subgraph sits inside
a complex expression. The collapse to a common mode happens at apply time
instead.

```{r no-rewrite}
Areal <- linop(matrix(1:4, 2, 2))
t(Areal)$node
adjoint(Areal)$node
all.equal(t(Areal) %*% c(1, 1), adjoint(Areal) %*% c(1, 1))
```

## Applying the tree

An apply carries a mode: `N` for `A X`, `T` for `A^T X`, `C` for `A^H X`, `R` for
`conj(A) X`. A view node does not do arithmetic. It composes its own mode with
the incoming one and passes the block down, which is why `t(adjoint(A))` costs
exactly what `A` costs.

The composition table is the Klein four-group, so two views always collapse to
one mode and nesting is free:

|  | N | T | C | R |
|---|---|---|---|---|
| **N** | N | T | C | R |
| **T** | T | N | R | C |
| **C** | C | R | N | T |
| **R** | R | C | T | N |

A product applies right to left under `N` and `R`, and left to right under `T`
and `C`, because `(A B)^T = B^T A^T`. A sum applies each term to the same block
and adds. Neither ever forms an intermediate operator, only intermediate blocks
the width of the one you passed in.

```{r apply-cost}
X <- matrix(rnorm(n * 3), n, 3)
dim(E %*% X)
```

## The cost model, and what it is not

`linop_cost()` reports estimated flops per column.

```{r cost}
linop_cost(S)
linop_cost(E)
```

Leaves declare their own cost, composites add up their children, and the number
is used to order candidate plans. It is not a prediction of wall time, and the
package is explicit about why. Timing during design found that a wider block is
*slower per column* for memory-bound operators: a pure diagonal scale at
n = 1e5 ran at 0.65x the per-column throughput at block width 64 compared with
width 1, because `d * X` recycled down a tall matrix leaves cache. Dense leaves
gained 2.2x over the same change, since a wide block turns BLAS-2 into BLAS-3.
A flop count cannot express either effect.

Read the cost as an ordering, and measure when the answer matters.

## Materialisation

Three things could reasonably be called forming a matrix, and the package treats
them differently. Building the operator you were handed is refused unless asked
for in as many words. Tall solver workspace of shape n by k is ordinary and
allowed. Small projected problems of shape k by k are ordinary and allowed.

The explicit request is `as.matrix()`, guarded on the entry count:

```{r as-matrix}
round(as.matrix(E), 3)
```

```{r as-matrix-guard, error = TRUE}
huge <- linop(function(X) X, adjoint = function(X) X, dim = c(50000, 50000))
as.matrix(huge)
```

The guard names the size in gigabytes and the argument that raises it, so
overriding it is a decision rather than an accident. `as_sparse()` does the same
through **Matrix**, returning the stored matrix untouched when the operator is
already a sparse leaf.

## collapse()

Between "never form anything" and "form the whole thing" sits the case where a
subexpression is applied many times and is small enough to hold. `collapse()`
walks the tree, replaces such subexpressions with dense leaves, and hands back a
report.

```{r collapse}
K <- linop(matrix(rnorm(200 * 200), 200, 200))
G <- crossprod(K) + 0.5 * linop_eye(200)

linop_cost(G)
Gc <- collapse(G, expected_applies = 500, max_entries = 1e6)
linop_cost(Gc)
Gc$node
```

```{r collapse-report}
report <- attr(Gc, "report")
class(report) <- "linop_collapse_report"
report
```

Two arguments control it. `expected_applies` is how many times you expect to use
the operator, and `max_entries` is the largest leaf `collapse()` may build. The
report states the memory it added and the number of applications at which that
memory pays for itself, so the decision is inspectable after the fact.

Raise `max_entries` and nothing else changes about the object: the result is a
`linop` with the same action, verified by the same suite.

```{r collapse-same}
x <- rnorm(200)
all.equal(as.vector(G %*% x), as.vector(Gc %*% x))
```

## Immutability

Every operation returns a new object and none mutates its operands. This is what
makes an operator safe to share between expressions, safe to hold in a closure,
and safe to apply twice and compare, which the conformance suite does under the
name `purity`.

```{r immutable}
A1 <- linop(matrix(1:4, 2, 2))
A2 <- 2 * A1
A1$args$a          # A1 is untouched; the scale lives in A2
A2$args$a
```

## Dtype promotion

An operator is `"double"` or `"complex"`, and no path downcasts. A real operator
applied to a complex block promotes its output, and a sum or product of a real
and a complex operator is complex.

```{r dtype}
Ar <- linop(matrix(c(1, 0, 0, 1), 2, 2))
Az <- linop(matrix(c(1i, 0, 0, 1i), 2, 2))
dtype(Ar)
dtype(Az)
dtype(Ar + Az)
dtype(Ar %*% Az)
Ar %*% c(1 + 2i, 3)
```

The one place this is worth spelling out is the sparse leaf. **Matrix** 1.7.5
declares the virtual complex classes but not the concrete ones, so a complex
sparse operator cannot be built at all. A real sparse operator applied to a
complex block still works: the block is split into real and imaginary parts,
applied separately, and recombined.

```{r sparse-complex, eval = requireNamespace("Matrix", quietly = TRUE)}
Sp <- linop(Matrix::sparseMatrix(i = 1:3, j = 1:3, x = c(2, 3, 4)))
Sp %*% c(1 + 1i, 1, 1)
```

## Where to next

- `vignette("capabilities")` covers what travels up the tree alongside the
  action: what is known about an operator and the argument behind it.
- `vignette("adapters")` covers adding a node type of your own to the registry
  listed at the top of this vignette.
- `vignette("solvers")` covers solving against a tree like this one, without
  materialising it.
