---
title: "Getting started with linop"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with linop}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(linop)
set.seed(1)
```

## What a linop is

A matrix in R is a rectangle of numbers. Many of the linear maps that turn up in
practice are described more directly by what they do than by the numbers they
would contain: a finite-difference stencil, a convolution, a projection onto a
subspace, the Jacobian of a simulation, a Gram matrix of a design you already
hold in another form. Writing out the entries is possible for all of them and
affordable for few.

A `linop` is that map as an object. You supply the action, `linop` supplies
everything a matrix would have given you.

```{r first}
n <- 8
S <- linop(
  function(X) rbind(X[-1, , drop = FALSE], 0),
  adjoint = function(X) rbind(0, X[-n, , drop = FALSE]),
  dim = c(n, n)
)
S
```

`S` shifts a vector up by one position and pads with a zero. Its conjugate
transpose shifts down. Two closures and a shape, and the object is complete.

```{r apply-vector}
v <- 1:8
S %*% v
```

A vector goes in and a vector comes out, as it would with a matrix. Handing it a
matrix gives a matrix, one column at a time:

```{r apply-block}
S %*% cbind(1:8, rep(1, 8))
```

## Building one

`linop()` is a single generic dispatching on its first argument, with methods
for the inputs below and a clear error for anything else.

A base matrix wraps the storage you already have:

```{r from-matrix}
M <- matrix(c(2, 1, 0,
              1, 3, 1,
              0, 1, 2), 3, 3)
A <- linop(M)
A
```

Notice what it says about itself. The dense leaf checks symmetry, diagonality and
triangularity exactly, on data it already holds, and records the outcome. That
check is `all.equal()` against the transpose, not a guess.

A `Matrix` from the **Matrix** package wraps sparse storage:

```{r from-sparse, eval = requireNamespace("Matrix", quietly = TRUE)}
Sp <- Matrix::sparseMatrix(i = c(1, 2, 3), j = c(1, 2, 3), x = c(2, 3, 4))
linop(Sp) %*% c(1, 1, 1)
```

A function is the callback form used above. `dim` is required, `adjoint` is
optional, and omitting the adjoint means the transposed modes raise an error
naming what is missing rather than returning something wrong.

```{r no-adjoint, error = TRUE}
Twice <- linop(function(X) 2 * X, dim = c(4, 4))
t(Twice) %*% rep(1, 4)
```

An object with no `linop()` method gets an error that names the method to write:

```{r no-method, error = TRUE}
linop(list(1, 2))
```

Two named constructors cover the operators that appear constantly inside
expressions:

```{r constructors}
linop_eye(4)
linop_scaling(c(2, 4, 8))
```

## The algebra is lazy

Every arithmetic operation returns a new operator. Nothing is applied, nothing is
allocated.

```{r lazy}
B <- crossprod(S) + 0.1 * linop_eye(n)
B
```

The tree printed underneath is the expression as written. `crossprod(S)` became a
product of a transpose view and the original leaf; the scaled identity became a
`scale` over `I`; the sum holds both. Applying `B` walks that tree right to left,
one block at a time.

```{r lazy-apply}
B %*% rep(1, n)
```

Everything R offers for matrices is routed through one `matrixOps` group method,
so all three products behave the same way:

```{r matrixops}
X <- matrix(rnorm(n * 2), n, 2)
all.equal(crossprod(S) %*% X, t(S) %*% (S %*% X))
all.equal(tcrossprod(S) %*% X, S %*% (t(S) %*% X))
```

Two operations are deliberately absent, because for matrices they mean something
this package cannot mean. Adding a scalar to an operator would have to guess
between `A + c` elementwise and `A + cI`, and `*` between two operators is the
elementwise product:

```{r not-defined, error = TRUE}
A + 1
```

```{r not-defined-2, error = TRUE}
A * A
```

## Transpose, adjoint and conjugate are three things

For a real operator, the transpose and the conjugate transpose have the same
action. They stay separate objects anyway, so that the expression you printed is
the expression you wrote, and so that a real subgraph inside a complex expression
keeps the distinction at the point where it starts to matter.

```{r views}
Rect <- linop(matrix(1:6, 2, 3))
t(Rect)$node
adjoint(Rect)$node
Conj(Rect)$node
```

Composing them collapses through the identities you would expect, at construction
time:

```{r view-identities}
t(t(Rect))$node              # involution
Conj(adjoint(Rect))$node     # adjoint then conjugate is the transpose
adjoint(t(Rect))$node        # transpose then adjoint is the conjugate
```

Internally these are four apply modes in the BLAS sense, composed through a Klein
four-group: `N` for `A X`, `T` for `A^T X`, `C` for `A^H X`, `R` for
`conj(A) X`. A callback leaf supplies two of them and the other two are derived
by conjugation, which is why nobody writes four functions.

## Complex operators, and one thing to know about crossprod

An operator is `"double"` or `"complex"`. No path silently drops an imaginary
part: a real operator applied to a complex block promotes the result.

```{r dtype}
dtype(A)
A %*% c(1 + 2i, 0, 0)
```

Base R's `crossprod(A)` is `A^T A`, the transpose and not the conjugate
transpose. For a complex operator that product is symmetric and is generally not
Hermitian. `linop` follows base R, because the promise of the package is that a
`linop` behaves like a matrix:

```{r crossprod-complex}
Z <- linop(matrix(c(1 + 1i, 2, 0, 3 - 2i), 2, 2))
cap(crossprod(Z), "symmetric")$value
cap(crossprod(Z), "hermitian")$value
```

The Hermitian Gram matrix is spelled out:

```{r gram}
cap(adjoint(Z) %*% Z, "hermitian")$value
```

## Looking at an expression

`print()` shows the tree. `explain()` adds the shape, dtype and cost of every
node:

```{r explain}
explain(B)
```

The cost column is flops per column, and it is an ordering heuristic. Measurement
during design found operators where a wider block is *slower* per column, which a
flop count cannot express, so this number orders candidate plans and never
predicts wall time.

`summary()` reports the capability table with the evidence behind each entry:

```{r summary}
summary(B)$caps[, c("capability", "value")]
```

## Forming a matrix, when you mean to

Materialisation is an explicit request with a guard, because the entire point of
the object is that it normally does not happen.

```{r materialize}
as.matrix(A)
round(as.matrix(B), 3)
```

The guard is on the entry count and refuses politely:

```{r guard, error = TRUE}
big <- linop(function(X) X, adjoint = function(X) X, dim = c(1e5, 1e5))
as.matrix(big)
```

`collapse()` is the middle path. It walks the expression and replaces
subexpressions that are cheaper to hold as a dense leaf, and it reports what it
did, so a wrong cost model costs a suboptimal plan and never a surprise
allocation.

```{r collapse}
K <- linop(matrix(rnorm(200 * 200), 200, 200))
E <- crossprod(K) + 0.5 * linop_eye(200)
linop_cost(E)

Ec <- collapse(E, expected_applies = 500)
linop_cost(Ec)
```

```{r collapse-report}
report <- attr(Ec, "report")
class(report) <- "linop_collapse_report"
report
```

## Checking that it is what it claims

`verify()` runs the contract. It is the function to reach for after writing a
hand-rolled operator, and the one an adapter author runs before shipping.

```{r verify}
verify(S)
```

The interesting line is the first. `adjoint consistency` tests
`<y, A x> = <A^H y, x>` over random probes, which is what catches a missing or
misplaced conjugation. Here is the same operator with its adjoint replaced by a
copy of the forward action:

```{r verify-bad}
Bad <- linop(function(X) rbind(X[-1, , drop = FALSE], 0),
             adjoint = function(X) rbind(X[-1, , drop = FALSE], 0),
             dim = c(6, 6))
verify(Bad)$overall
```

The full certificate, and what each check is for, is covered in
`vignette("adapters")`.

## Solving

An operator you can apply is an operator you can solve against. `solve()` is R's
own generic, so it needs no new verb:

```{r solve}
Lap <- linop(function(X) 2 * X - rbind(X[-1, , drop = FALSE], 0) -
                         rbind(0, X[-8, , drop = FALSE]), dim = c(8, 8))
rhs <- as.numeric(Lap %*% rep(1, 8))
round(solve(Lap, rhs), 12)
```

`eigs()` and `svds()` are the other two, for eigenpairs and singular triplets.
All three return a certificate alongside the answer.

## Where to next

- `vignette("expressions")` covers the expression graph: the simplification
  rules, the cost model, `collapse()` and materialisation.
- `vignette("capabilities")` covers what an operator knows about itself and how
  a solver states what it will accept.
- `vignette("solvers")` covers `solve()` and the seven Krylov methods behind it,
  with preconditioners.
- `vignette("spectral")` covers `eigs()` and `svds()`.
- `vignette("certificates")` covers what comes back attached to every answer.
- `vignette("adapters")` covers writing `linop.myclass()`, registering a node
  type, and the preconditioner contract.
