---
title: "Operators on a sequence space"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Operators on a sequence space}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(linop)
```

## An operator with no matrix and no size

A `linop` is defined by its action rather than its entries, and a dimension of
`Inf` is what that promise looks like taken all the way. `linop_jacobi()` builds
a self-adjoint operator on `l^2(Z)`, the square-summable sequences indexed by the
integers:

```{r build}
H <- linop_jacobi(diagonal = 1)
H
dim(H)
```

It is tridiagonal in the standard basis:

`(H u)[j] = u[j-1] + u[j+1] + V[j] u[j]`

with `V` here the potential `1` at the origin and zero elsewhere. Both coefficient
sequences are real and *eventually constant*: they take the values you supply on a
finite window and their limiting values everywhere else.

Nothing numeric runs on it, and the refusals say so by name:

```{r refuse, error = TRUE}
as.matrix(H)
```

`Inf` is the extent of the index set here, not a flag. The ordinary shape
arithmetic keeps working on it with no special case, so the algebra composes
these operators as it composes any other:

```{r algebra}
dim(t(H))
dim(2 * H + H)
```

## Truncating it

`finite_section(H, n)` is `P_n H P_n` on the indices `-n` to `n`. What comes back
is an ordinary operator of dimension `2n + 1`:

```{r section}
F <- finite_section(H, n = 20)
F
dim(F)
```

The printed tree shows the infinite operator underneath the finite one, because
that is what the section holds. It is not an annotation: the child is the operator
itself, and everything the certificate says about it is read off that child.

```{r structure}
explain(F)
```

A section is an ordinary `linop` in every other respect. It has a matrix, the
solvers run on it, and it composes:

```{r ordinary}
round(as.matrix(finite_section(H, n = 2)), 3)
x <- solve(F, rep(1, 41), method = "minres")
max(abs(as.numeric(F %*% x) - 1))
```

`n` has to be large enough that the window sits strictly inside it. Outside the
window the coefficients are at their limits and the eigenvalue equation is the
free recurrence, which is what makes everything below a closed form:

```{r narrow, error = TRUE}
finite_section(linop_jacobi(diagonal = c(1, 2, 3)), n = 2)
```

## Two things that are known exactly

The reason this class comes first is that two quantities are exact rather than
estimated.

The perturbation is finitely supported, hence finite rank, hence compact, so by
Weyl's theorem the essential spectrum is that of the free operator. The free
operator has symbol `a + 2 b cos(theta)`, so:

`sigma_ess(H) = [a - 2|b|, a + 2|b|]`

which for this operator is `[-2, 2]`, from the coefficients and with nothing
computed.

The second is the decay. Outside the window an eigenfunction satisfies the free
recurrence exactly, whose solutions are `z^j` for the two roots of
`z^2 - mu z + 1` with `mu = (lambda - a)/b`. The roots are reciprocal, so for a
value outside the band exactly one is inside the unit disc, and square
summability forces it on each side:

```{r decay}
decay_rate(H, sqrt(5))
(sqrt(5) - 1) / 2
```

The eigenfunction is *exactly* geometric outside the window at that rate, not
asymptotically so. Inside the band both roots sit on the unit circle, nothing
decays, and there is nothing to report:

```{r decay-band}
decay_rate(H, c(0, 1.9, 2, 3))
```

## Solving it

`eigs()` on the section certifies a statement about `H`:

```{r eigs}
fit <- eigs(F, k = 1, which = "largest_algebraic")
fit$values
```

For a single site at the origin the answer is closed form,
`lambda = sign(v) sqrt(v^2 + 4)`, so the error is available:

```{r truth}
c(computed = fit$values, exact = sqrt(5), error = abs(fit$values - sqrt(5)))
```

and the certificate bounded it without knowing it:

```{r bound}
v <- certificate(fit)$values
c(bound = v$bound, true_error = abs(fit$values - sqrt(5)))
```

`certificate()` is the reader. Every verb in the package returns one somewhere in
its result, and this reaches it wherever that is.

That bound is two terms in quadrature and a floor. The truncation term is two
entries of the computed eigenvector and nothing else, and it falls at the closed-
form rate:

```{r ladder}
rho <- decay_rate(H, sqrt(5))
t(vapply(c(10, 20, 30, 40), function(n) {
  cv <- eigs(finite_section(H, n = n), k = 1,
             which = "largest_algebraic")$certificate$values
  c(n = n, truncation = cv$truncation, over_rho_n = cv$truncation / rho^n)
}, numeric(3)))
```

The last column is constant, which says the rate is `rho` with a computable
constant rather than merely `O(rho^n)`. Push far enough and it stops being
constant, because the tail has fallen below what the finite eigensolve resolves
and the term is reading noise rather than truncation:

```{r plateau}
deep <- eigs(finite_section(H, n = 80), k = 1, which = "largest_algebraic")
c(measured = deep$certificate$values$truncation, analytic = 0.5845 * rho^80)
```

The bound still holds, because it is an identity about the vector actually
computed rather than a statement about the exact eigenvector. Where the term has
gone all the way down to the arithmetic floor the row says so:

```{r plateau-floor}
sharp <- eigs(finite_section(linop_jacobi(diagonal = 4), n = 40), k = 1,
              which = "largest_algebraic")
ch <- sharp$certificate$checks
cat(ch$detail[ch$check == "truncation bound"])
```

## Letting it choose the width

`n = 20` above was a guess. `eigs()` on the operator itself does not need one:

```{r direct}
chosen <- eigs(H, k = 1, which = "largest_algebraic")
chosen
```

No width is available before the solve: the tail falls at
`decay_rate(H, lambda)`, and `lambda` is what the solve produces. So the width is
measured rather than derived, and one measurement is enough. Outside the window
the eigenvector is exactly geometric, so the truncation term at one width gives it
at every other,

`eta(n') = eta(n) rho^(n' - n)`,

and the width that brings it to a target is a division:

`n' = n + log(target / eta(n)) / log(rho)`

The widths it went through are on the result, and there are two of them: one to
measure and one to confirm.

```{r widths}
chosen$widths
```

That holds at every rate. A potential nearer the band edge decays more slowly and
wants a wider section, and the number of widths stays the same:

```{r rates}
t(vapply(c(0.2, 0.3, 0.5, 1, 2), function(v) {
  Hv <- linop_jacobi(diagonal = v)
  f <- eigs(Hv, k = 1, which = "largest_algebraic")
  c(v = v, rho = decay_rate(Hv, f$values), n = f$n, widths = length(f$widths))
}, numeric(4)))
```

The last row took one, because the first section was already wide enough at
`v = 2`. The rows above it are what the closed form buys: from a first width of
41 a doubling scheme reaches 82 and then 164, and `v = 0.3` wants 150.

What stopped the search is on the certificate, where every other dispatch in the
package is recorded:

```{r stopped}
certificate(chosen)
```

Three things stop it. The tail can reach the target, which is the line above. It
can stop falling with `n`, which is the plateau: past the level at which the
finite eigensolve stores the tail at all, `eta` measures the eigenvector's own
error and no width improves it. And `n_max` is a budget, reported as the width it
would have taken:

```{r capped}
capped <- eigs(linop_jacobi(diagonal = 0.3), k = 1, which = "largest_algebraic",
               section = list(n_max = 60))
capped$certificate$dispatch$reason
ch <- certificate(capped)$checks
ch[ch$check %in% c("finite residual", "truncation bound"), c("check", "status")]
```

The eigensolve was fine and the section was short, and the two rows say so
separately, because they answer to different knobs.

## The subspace is the knob to watch

A section's spectrum is a discretisation of the essential spectrum together with
whatever separated from it. Enlarging `n` densifies that cluster rather than
adding isolated eigenvalues, so a value close to the band edge has to be told
apart from an ever tighter crowd. `eigs()` widens its default subspace for a
section because of it, and a narrower one fails in a way worth seeing:

```{r ncv}
W <- finite_section(linop_jacobi(diagonal = 0.3), n = 80)
narrow <- eigs(W, k = 1, which = "largest_algebraic", ncv = 21)
wide   <- eigs(W, k = 1, which = "largest_algebraic")

c(narrow = narrow$values, wide = wide$values, exact = sqrt(0.3^2 + 4))
```

The narrow run is wrong in the second digit, and it reports that it is. The row
to read is the eigensolve's own residual, which is eleven orders apart:

```{r ncv-cert}
c(narrow = narrow$certificate$values$residual,
  wide   = wide$certificate$values$residual)
```

More budget does not recover it and a wider section makes it worse, which is what
separates this from an ordinary shortage of iterations. See
`vignette("spectral")` for `ncv` on operators generally.

## When there is nothing to find

A potential of zero leaves the free operator, which has no eigenvalues at all.
Every finite section still produces numbers, and they look like eigenvalues:

```{r free}
free <- eigs(finite_section(linop_jacobi(diagonal = 0), n = 40), k = 1,
             which = "largest_algebraic")
free$values
free$certificate$values$residual
```

The value sits just inside `[-2, 2]` and approaches the edge from below as `n`
grows. The eigensolve converged cleanly and has nothing to report. What does is
the isolation row, which tests the value against the essential spectrum:

```{r free-cert}
free$certificate
```

This is the case the exact essential spectrum exists for. A value that has not
cleared the band is a discretisation of continuous spectrum, and the certificate
refuses rather than issuing a number that would read like an eigenvalue.

The width search meets the same wall. With nothing outside the band there is no
rate to predict with, so it doubles until its budget and reports that:

```{r free-search}
hunt <- eigs(linop_jacobi(diagonal = 0), k = 1, which = "largest_algebraic",
             section = list(n_max = 200))
hunt$widths
hunt$certificate$dispatch$reason
```

## Where to next

- `vignette("certificates")` covers the fifth certificate shape beside the other
  four.
- `vignette("spectral")` covers `eigs()` and `svds()` on ordinary operators.
- `vignette("expressions")` covers the algebra these operators compose under.
