---
title: "Solving a system"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Solving a system}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(linop)
set.seed(1)
```

## One verb

`solve()` is R's generic for `a x = b`, and a `linop` has a method for it. There
is no second verb to learn and no separate object to construct.

```{r first}
n <- 200
L <- linop(
  function(X) 2 * X - rbind(X[-1, , drop = FALSE], 0) - rbind(0, X[-n, , drop = FALSE]),
  dim = c(n, n)
)
xstar <- sin(seq_len(n) * pi / (n + 1))
b <- as.numeric(L %*% xstar)

x <- solve(L, b)
max(abs(x - xstar))
```

`L` is the 1D Laplacian written as its action. Nothing was factorized, because
for an operator given as a closure there is nothing to factorize. What came back
is a plain numeric vector, so it indexes and multiplies like any other, and it
carries a record of how it was produced:

```{r cert}
attr(x, "certificate")
```

The first line is the one to read here. The rest of that table is the subject of
`vignette("certificates")`.

## What `auto` chose, and why

`method = "auto"` picks from what the operator has established about itself and
from its shape, and records both the choice and the reason. Three routes are
reachable.

A diagonal operator of full rank is solved exactly, with no iteration at all:

```{r auto-direct}
D <- linop_scaling(seq(1, 4, length.out = 50))
attr(solve(D, rep(1, 50)), "certificate")$dispatch$reason
```

A dense leaf checks its own symmetry when it is built, on data it already holds,
so a symmetric matrix reaches MINRES:

```{r auto-minres}
M <- crossprod(matrix(rnorm(60 * 60), 60, 60)) + diag(60)
Ad <- linop(M)
bd <- as.numeric(Ad %*% rep(1, 60))
attr(solve(Ad, bd), "certificate")$dispatch$reason
```

The operator above is positive definite, and `auto` still chose MINRES rather
than CG. Symmetry was computed; definiteness was not, and `auto` will not assume
it. There is a CG branch above the MINRES one, and no constructor in the package
currently reaches it, because the routes that establish definiteness at a source
`auto` accepts are diagonal and are taken by the direct route first. CG is
reached by naming it.

That is the general rule, and it applies to declarations too. `L` at the top of
this article was built with no properties at all. Declaring them changes nothing
about which method `auto` picks:

```{r auto-declared}
Ldecl <- linop(
  function(X) 2 * X - rbind(X[-1, , drop = FALSE], 0) - rbind(0, X[-n, , drop = FALSE]),
  dim = c(n, n),
  properties = c(hermitian = TRUE, positive_definite = TRUE)
)
attr(solve(Ldecl, b), "certificate")$dispatch$reason
```

Both capabilities are `TRUE` and `auto` acted on neither, because a bare
declaration is not the kind of argument `auto` will choose a method on. The
capability and the evidence behind it are two separate questions, which is what
`vignette("capabilities")` is about. Here it decides a dispatch.

GMRES is the fallback, and it is what makes `auto` total rather than partial: it
requires nothing of the operator, and it runs on the forward action alone. `L`
has no adjoint, which is why every route above it was unavailable in the first
place.

## Naming a method

Naming a method is the caller making the assertion, so no evidence minimum
applies:

```{r named}
attr(solve(Ldecl, b, method = "cg"), "certificate")$dispatch
```

The declaration on `Ldecl` is true, CG is the right method for it, and saying so
is how you get it. The two questions stay separate: `auto` is the package
choosing on your behalf, and a named method is your own claim about your own
operator.

A declaration that is false is not silently tolerated. CG forms `p^H A p` every
step, and a value that is not positive far from convergence means the operator is
not what it said it was, which is an error naming the capability rather than a
wrong answer.

## The seven methods

```{r roster, echo = FALSE}
knitr::kable(data.frame(
  method = c("cg", "minres", "gmres", "fgmres", "bicgstab", "lsqr", "lsmr"),
  requires = c("hermitian positive definite", "hermitian", "nothing", "nothing",
               "nothing", "nothing", "nothing"),
  adjoint = c("no", "no", "no", "no", "no", "yes", "yes"),
  shape = c("square", "square", "square", "square", "square",
            "any", "any"),
  check.names = FALSE))
```

GMRES and BiCGSTAB both require nothing and both run without an adjoint. They are
not interchangeable. GMRES minimises the residual over the whole Krylov space, so
its residual cannot rise and it has no breakdown; BiCGSTAB is cheaper per step
and can break down, in which case it returns a `fail` certificate naming the
breakdown rather than throwing. `auto` prefers GMRES for the first reason.

FGMRES is GMRES with the one difference that it accepts a preconditioner that
changes between applications, such as an inner iterative solve.

## Least squares is a different request

`solve()` with `method = "auto"` refuses a rectangular operator, and says what to
name instead:

```{r rect-refuse, error = TRUE}
G <- matrix(rnorm(120 * 40), 120, 40)
R <- linop(G)
bl <- rnorm(120)
solve(R, bl)
```

This is deliberate. `min ||b - A x||` is a different mathematical question from
`A x = b`, rather than the same one on a different shape, so the caller states
which they meant. Both methods are reachable through the same verb:

```{r lsqr}
xl <- solve(R, bl, method = "lsqr")
max(abs(xl - qr.solve(G, bl)))
```

LSQR and LSMR solve the same problem and minimise different things over the same
bidiagonalisation. LSQR minimises `||r||`. LSMR minimises `||A^H r||`, which is
the quantity that goes to zero at a least-squares solution and the numerator of
the backward error the certificate reports:

```{r lsmr}
xm <- solve(R, bl, method = "lsmr")
c(lsqr = attr(xl, "certificate")$values$backward_error,
  lsmr = attr(xm, "certificate")$values$backward_error)
```

## Preconditioners

A preconditioner is built from one direction, `M^-1`, and carries what it
declares about itself:

```{r precond-build}
set.seed(2)
d <- 10^runif(200, 0, 5)
Adense <- diag(d)
Adense[cbind(1:199, 2:200)] <- -1
Adense[cbind(2:200, 1:199)] <- -1
A <- linop(Adense)
xs <- rep(1, 200)
bb <- as.numeric(A %*% xs)

P <- preconditioner(function(R) R / d, hermitian = TRUE,
                    positive_definite = TRUE, side = "left", dim = c(200, 200))
P
```

The diagonal of that operator spans five orders of magnitude. At a fixed budget
of 40 iterations, with everything else equal:

```{r precond-effect}
plain  <- solve(A, bb, method = "minres", maxit = 40, tol = 1e-10)
scaled <- solve(A, bb, method = "minres", maxit = 40, tol = 1e-10,
                preconditioner = P)

c(plain  = max(abs(plain  - xs)),
  scaled = max(abs(scaled - xs)))
c(plain  = attr(plain,  "certificate")$overall,
  scaled = attr(scaled, "certificate")$overall)
```

Each method states which properties it needs of a preconditioner and which sides
it is defined for, and enforces both at the point the preconditioner is accepted:

```{r precond-table, echo = FALSE}
knitr::kable(data.frame(
  method = c("cg", "minres", "gmres", "fgmres", "bicgstab", "lsqr", "lsmr"),
  requires = c("fixed, hermitian, positive definite",
               "fixed, hermitian, positive definite",
               "fixed", "nothing", "fixed", "fixed", "fixed"),
  sides = c("left, right, split", "left, right, split", "left, right, split",
            "right", "left, right, split", "right", "right"),
  check.names = FALSE))
```

A refusal names what was missing:

```{r precond-refuse, error = TRUE}
flexible <- preconditioner(function(R) R / d, fixed = FALSE,
                           hermitian = TRUE, positive_definite = TRUE,
                           side = "left", dim = c(200, 200))
solve(A, bb, method = "minres", preconditioner = flexible)
```

For CG the three sides coincide, so its row is unrestricted because the
distinction does not arise. For GMRES all three are implemented and they produce
different iterates, so its row is unrestricted for the opposite reason:

```{r sides}
set.seed(1)
Gm <- matrix(rnorm(80 * 80), 80, 80)
diag(Gm) <- diag(Gm) + 12
Ag <- linop(Gm)
bg <- as.numeric(Ag %*% rep(1, 80))
dg <- diag(Gm)

vapply(c("left", "right", "split"), function(sd) {
  Ps <- preconditioner(function(R) R / dg, hermitian = TRUE,
                       positive_definite = TRUE, side = sd, dim = c(80, 80))
  attr(solve(Ag, bg, method = "gmres", maxit = 6, tol = 1e-12,
             preconditioner = Ps), "certificate")$values$residual
}, numeric(1))
```

Right preconditioning minimises the residual that gets reported. Left and split
minimise a preconditioned one, so the outer loop converts the target going in and
measures in your units coming out.

The narrow rows are the two least-squares methods, and the reason is the shape.
`M^-1` acts on the domain of the operator; left and split act on the codomain,
which for a rectangular operator is a different space.

## A solve that does not converge

An exhausted budget is a result, not an exception. It comes back as a `fail`
certificate that says what ran out:

```{r nonconv}
attr(plain, "certificate")
```

The iterate is finite and usable, the number reached is reported, and nothing was
thrown. A contradicted declaration is the case that does throw, because there the
operator is not what it claimed and no iterate is worth returning.

`details = TRUE` returns the whole object rather than the solution, including the
convergence history:

```{r details}
fit <- solve(A, bb, method = "minres", maxit = 40, tol = 1e-10, details = TRUE)
names(fit)
fit$iterations
fit$converged
```

## Where to next

- `vignette("certificates")` covers the table printed above: what each row
  measures, the arithmetic floor underneath it, and which rows carry a bound
  rather than a measurement.
- `vignette("spectral")` covers `eigs()` and `svds()`, which are built on the
  solvers here when a shift is given.
- `vignette("capabilities")` covers what an operator establishes about itself,
  which is what `auto` reads.
