---
title: "Eigenvalues and singular values"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Eigenvalues and singular values}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(linop)
set.seed(1)
```

## Two verbs

`eigs()` computes a few eigenpairs of a hermitian operator. `svds()` computes a
few singular triplets of any operator. Both run against the action alone, and
both hold `ncv` vectors rather than a matrix.

The 1D Laplacian has closed-form eigenvalues, which makes it the fixture to check
against:

```{r first}
n <- 200
lap <- function(X) 2 * X - rbind(X[-1, , drop = FALSE], 0) - rbind(0, X[-n, , drop = FALSE])
L <- linop(lap, adjoint = lap, dim = c(n, n), properties = c(hermitian = TRUE))

truth <- 2 - 2 * cos(seq_len(n) * pi / (n + 1))

fit <- eigs(L, k = 3, which = "largest_algebraic", ncv = 60)
fit
```

```{r first-check}
rbind(computed = fit$values,
      closed_form = rev(sort(truth))[1:3])
```

## The operator has to say it is hermitian

There is one eigensolver here and it is Lanczos, which is defined for a hermitian
operator. Symmetry is not probed and not assumed:

```{r no-declaration, error = TRUE}
Undeclared <- linop(lap, adjoint = lap, dim = c(n, n))
eigs(Undeclared, k = 2)
```

Unknown is not false, and the refusal says so rather than guessing either way. A
dense leaf establishes symmetry by checking it, so a matrix needs no declaration;
a callback has nothing to check and the declaration is how you supply it.

A non-hermitian operator is refused for a different reason, and there is no
Arnoldi in this version, so it has no eigensolver here at all. `svds()` works on
any operator and is often the thing that was actually wanted.

## `ncv` is the knob that matters

`ncv` is how many vectors a restart holds. It is the binding constraint on
whether the method converges, more than the tolerance and more than the budget:

```{r ncv-ladder}
ladder <- t(vapply(c(20, 30, 40, 60), function(nc) {
  f <- eigs(L, k = 3, which = "largest_algebraic", ncv = nc)
  c(ncv = nc,
    error = max(abs(f$values - rev(sort(truth))[1:3])),
    converged = f$nconv)
}, numeric(3)))
ladder
```

At `ncv = 20` the answer is wrong in the third digit and nothing converges. At
`ncv = 60` every pair is right to the last one. Raising the iteration budget does
not substitute: the subspace stops improving, the method reports that it stopped
improving, and it returns what it has.

The top of the Laplacian's spectrum is tightly clustered, which is what makes a
small subspace insufficient here. A well-separated spectrum needs far less.

## What gets reported is the Rayleigh quotient

The value attached to each vector is `x^H A x / (x^H x)`, measured against `A`
itself rather than read off the projected problem. It is the choice of `mu` that
minimises `||A x - mu x||`, so no other value certifies better for the vector
that was found.

That matters most under a shift, where the projected problem is a solve against
`A - sigma I` and its own accuracy is a separate question. Measuring on `A`
means nothing downstream inherits the inner solve's error.

## Which end

```{r which}
c(largest  = eigs(L, k = 2, which = "largest_algebraic",  ncv = 60)$values,
  smallest = eigs(L, k = 2, which = "smallest_algebraic", ncv = 40)$values)
```

`"largest"` and `"smallest"` order by magnitude, `"largest_algebraic"` and
`"smallest_algebraic"` by value. For an operator with a negative eigenvalue those
are different questions. RSpectra's `"LM"`, `"SM"`, `"LA"` and `"SA"` are
accepted for the same four, so existing calling code ports without edits.

## Eigenvalues near a point

A shift is the way to reach the interior of a spectrum. `sigma` runs the same
recurrence on `(A - sigma I)^-1`, and that inverse is not formed: it is the
package's own solvers, applied once per step.

```{r shift}
near <- eigs(L, k = 3, sigma = 0.01, ncv = 30, inner = list(method = "minres"))
rbind(computed = sort(near$values),
      closed_form = sort(truth[order(abs(truth - 0.01))][1:3]))
```

`inner` carries the arguments for that solve, so its method and tolerance are
yours to set. This is the one place the two halves of the package meet:
everything in `vignette("solvers")` is available inside a spectral computation.

A shift is expensive, because each Lanczos step now costs a full linear solve.
It is the right tool when the eigenvalues wanted are in the middle, where no
amount of `ncv` will reach them directly.

## Singular values

`svds()` needs both the forward action and the adjoint, and requires nothing else:

```{r svds}
G <- matrix(rnorm(60 * 12), 60, 12)
R <- linop(G)

sv <- svds(R, k = 2)
rbind(computed = sv$d, closed_form = svd(G)$d[1:2])
```

The triplets satisfy what they should:

```{r svds-check}
c(`A v - d u` = max(abs(G %*% sv$v - sv$u %*% diag(sv$d))),
  `U orthonormal` = max(abs(crossprod(sv$u) - diag(2))),
  `V orthonormal` = max(abs(crossprod(sv$v) - diag(2))))
```

Underneath, this is `eigs()` on the hermitian augmented operator

```
H = [ 0    A  ]
    [ A^H  0  ]
```

whose eigenpairs are `(sigma, [u; v]/sqrt(2))`. `H` is hermitian for every `A`
with nothing to declare, which is why `svds()` asks the operator for no
properties where `eigs()` asks for one. One consequence is worth knowing: the
certificate's `orthogonality` row measures the augmented basis, where a deviation
in `U` could cancel one in `V`, so the check above measures the two separately.

`which = "smallest"` is accepted and is the hard end of the problem. A Krylov
space reaches the largest singular values first, so the smallest need a subspace
approaching the rank of the operator:

```{r svds-smallest}
small <- svds(R, k = 2, which = "smallest", ncv = 12)
rbind(computed = small$d, closed_form = rev(svd(G)$d)[1:2])
```

The request is not refused. What was reached is reported, and the certificate is
where you find out whether it was enough.

## A result that is right and certifies as `fail`

This is worth meeting once, because it looks like a bug and is not:

```{r fail-but-right}
smallest <- eigs(L, k = 3, which = "smallest_algebraic", ncv = 40)

rbind(computed = smallest$values, closed_form = sort(truth)[1:3])
smallest$certificate$overall
```

Every value is right to the last printed digit and the certificate says `fail`.
The reason is in which row failed:

```{r fail-rows}
smallest$certificate
```

`backward error` passes at 1.4e-13 and `forward error`, which is a real bound
here, passes at 5.2e-13. What failed is `residual`, which divides by `|theta|`,
and the smallest eigenvalue of this operator is 2.4e-4. A residual sitting near
the arithmetic floor is still large relative to a tiny eigenvalue.

`overall` is a summary of seven rows, and each row answers its own question. The
rows are what to read. `vignette("certificates")` covers what each one measures.

## Where to next

- `vignette("certificates")` covers the table above, including why
  `forward error` carries a genuine bound for an eigenpair and `not_checked` for
  a linear solve.
- `vignette("solvers")` covers the linear solvers, which `sigma` runs on.
